---
title: "Expansion Teams"
format: 
  html:
    fig-width: 10
    fig-height: 6
    toc: false
lightbox: true
---

In 2025, somebody on Reddit asked me why the Valkyries initial Elo Rating of 1300 was so low, seemingly putting them at a disadvantage in the rankings.
While I didn't choose the value, 1300 is a default that has generally been appropriate for expansion teams, whose average end-of-season rating is 1306 (or 1291 if you exclude the Valks).
The Valkyries just happened to be a historically good expansion team.
To illustrate that point, I decided to compare all the expansion teams throughout the history of the league:

```{r}
#| include: false
source('code/setup.R')
source('code/utils.R')
source('code/plotting.R')

expanYrs <- read.csv(text = 'team,year
Detroit Shock,1998
Washington Mystics,1998
Minnesota Lynx,1999
Orlando Miracle,1999
Indiana Fever,2000
Miami Sol,2000
Portland Fire,2000
Seattle Storm,2000
Chicago Sky,2006
Atlanta Dream,2008
Golden State Valkyries,2025')

getEloCurve <- function(team, year, allgames){
    
    games <- allgames[which(allgames$season == year), ]
    
    # exclude playoffs, include CC
    games <- games[seq_len(max(which(!games$playoffs))), ]
    dateRange <- range(games$Date)
    
    # get elo from home and away games
    eloA <- games[games$`Visitor/Neutral` == team, c('Date','Visitor/Neutral','season','away_abbr','away_elo_pre','away_elo_post')]
    eloH <- games[games$`Home/Neutral` == team, c('Date','Home/Neutral','season','home_abbr','home_elo_pre','home_elo_post')]
    names(eloA) <- c('Date','Team_Name','season','team_abbr','elo_pre','elo_post')
    names(eloH) <- names(eloA)
    elo <- rbind(eloA, eloH)
    elo <- elo[order(elo$Date), c('Date','elo_pre','elo_post')]
    rm(eloA,eloH)
    # make elo curve by setting elo_pre as the previous day
    elopre <- elo[, c('Date','elo_pre')]
    elopre$Date <- elopre$Date - 1
    elopost <- elo[, c('Date','elo_post')]
    names(elopre) <- names(elopost) <- c('Date','elo')
    # there can be back-to-back games, so don't include elo_pre if there was a game that day
    elopre <- elopre[!elopre$Date %in% elopost$Date, ]
    elo <- rbind(elopre,elopost)
    elo <- elo[order(elo$Date), ]
    
    # add beginning and end
    if(! dateRange[1] %in% elo$Date){
        elo <- rbind(elo[1,], elo)
        elo$Date[1] <- dateRange[1]
    }
    if(! dateRange[2] %in% elo$Date){
        elo <- rbind(elo, elo[nrow(elo),])
        elo$Date[nrow(elo)] <- dateRange[2]
    }
    return(elo)
}

```


```{r}
#| echo: false
#| message: false
# merge with team abbreviations and color data
expanYrs <- merge(expanYrs, teamdata, all.x = TRUE)

# get elo curves as percentage of season
curves <- lapply(seq_len(nrow(expanYrs)), function(i){
    cv <- getEloCurve(expanYrs$team[i], expanYrs$year[i], allgames)
    cv$Date <- as.numeric(cv$Date)
    cv$Date <- 100 * (cv$Date - min(cv$Date)) / (max(cv$Date) - min(cv$Date))
    return(cv)
})

# sort by final elo
ord <- order(sapply(curves, function(x){ x$elo[nrow(x)] }))
curves <- curves[ord]
expanYrs <- expanYrs[ord, ]
expanYrs$elo <- sapply(curves, function(x){ x$elo[nrow(x)] })

# build legend df
lgnd <- expanYrs[nrow(expanYrs):1, ]
lgnd$label <- paste0(lgnd$abbr,' ', lgnd$year,' (',round(lgnd$elo),')')
```



```{r}
#| echo: false
#| message: false
#| classes: light-mode

layout(matrix(c(1,1,1,2), nrow=1))
par(mar = c(5,5,4,0)+.1, cex.lab = 1.25, cex.main = 2, bg = '#fffaf6')
# if(mode == 'dark'){
#     par(fg = 'white', bg = '#272935', col.axis = 'white', col.lab = 'white', col.main = 'white', col.sub = 'white')
# }

require(scales)
xl <- range(do.call(rbind, curves)$Date)
yl <- range(sapply(curves,function(x){x$elo}))
plot(xl, yl, col = 'transparent',
     xlab = 'Date (% of Season)', ylab='Elo Rating', main = 'Expansion Team Elo Ratings', las = 1)
abline(h = c(1300,1500), lty = 2)

for(i in 1:length(curves)){
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color1[i], lwd=4)
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color2[i], lwd=1)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color2[i], pch=16, cex=2)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color1[i], pch=16, cex = 1.5)
}

par(mar = c(5,0,4,1)+.1)
plot.new()
legend('left', legend = rep('', nrow(lgnd)), bty='n', lwd=4.5, 
       col = lgnd$color1, cex = 1.3)
legend('left', legend = lgnd$label, bty='n', lwd=1.1, 
       col = lgnd$color2, cex = 1.3)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode

layout(matrix(c(1,1,1,2), nrow=1))
par(mar = c(5,5,4,0)+.1, cex.lab = 1.25, cex.main = 2, bg = '#fffaf6')
par(fg = 'white', bg = '#272935', col.axis = 'white', col.lab = 'white', col.main = 'white', col.sub = 'white')

require(scales)
xl <- range(do.call(rbind, curves)$Date)
yl <- range(sapply(curves,function(x){x$elo}))
plot(xl, yl, col = 'transparent',
     xlab = 'Date (% of Season)', ylab='Elo Rating', main = 'Expansion Team Elo Ratings', las = 1)
abline(h = c(1300,1500), lty = 2)

for(i in 1:length(curves)){
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color1[i], lwd=4)
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color2[i], lwd=1)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color2[i], pch=16, cex=2)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color1[i], pch=16, cex = 1.5)
}

par(mar = c(5,0,4,1)+.1)
plot.new()
legend('left', legend = rep('', nrow(lgnd)), bty='n', lwd=4.5, 
       col = lgnd$color1, cex = 1.3)
legend('left', legend = lgnd$label, bty='n', lwd=1.1, 
       col = lgnd$color2, cex = 1.3)
```

The Valkyries' 2025 season stands above every other expansion team, with only the 1998 Detroit Shock coming close (in only the second year of the league's existence).
The two most recent expansion teams before the Valkyries were the Atlanta Dream (2008) and Chicago Sky (2006), both of whom finished well below the initial mark of 1300.



We can also look at these rankings by game number, to see if the Valkyries benefitted from an expanded schedule:

```{r}
#| echo: false
#| message: false
# merge with team abbreviations and color data
expanYrs <- merge(expanYrs, teamdata, all.x = TRUE)

# get elo curves as number of games
curves <- lapply(seq_len(nrow(expanYrs)), function(i){
    cv <- getEloCurve(expanYrs$team[i], expanYrs$year[i], allgames)
    keep <- c(TRUE, cv$elo[-1] != cv$elo[-length(cv$elo)])
    cv <- cv[keep, ]
    cv$Date <- c(0,seq_len(nrow(cv)-1))
    return(cv)
})

# sort by final elo
ord <- order(sapply(curves, function(x){ x$elo[nrow(x)] }))
curves <- curves[ord]
expanYrs <- expanYrs[ord, ]
expanYrs$elo <- sapply(curves, function(x){ x$elo[nrow(x)] })

# build legend df
lgnd <- expanYrs[nrow(expanYrs):1, ]
lgnd$label <- paste0(lgnd$abbr,' ', lgnd$year,' (',round(lgnd$elo),')')
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode

layout(matrix(c(1,1,1,2), nrow=1))
par(mar = c(5,5,4,0)+.1, cex.lab = 1.25, cex.main = 2, bg = '#fffaf6')
# if(mode == 'dark'){
#     par(fg = 'white', bg = '#272935', col.axis = 'white', col.lab = 'white', col.main = 'white', col.sub = 'white')
# }

require(scales)
xl <- range(do.call(rbind, curves)$Date)
yl <- range(sapply(curves,function(x){x$elo}))
plot(xl, yl, col = 'transparent',
     xlab = 'Game No.', ylab='Elo Rating', main = 'Expansion Team Elo Ratings', las = 1)
abline(h = c(1300,1500), lty = 2)

for(i in 1:length(curves)){
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color1[i], lwd=4)
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color2[i], lwd=1)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color2[i], pch=16, cex=2)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color1[i], pch=16, cex = 1.5)
}

par(mar = c(5,0,4,1)+.1)
plot.new()
legend('left', legend = rep('', nrow(lgnd)), bty='n', lwd=4.5, 
       col = lgnd$color1, cex = 1.3)
legend('left', legend = lgnd$label, bty='n', lwd=1.1, 
       col = lgnd$color2, cex = 1.3)
```
```{r}
#| echo: false
#| message: false
#| classes: dark-mode

layout(matrix(c(1,1,1,2), nrow=1))
par(mar = c(5,5,4,0)+.1, cex.lab = 1.25, cex.main = 2, bg = '#fffaf6')
par(fg = 'white', bg = '#272935', col.axis = 'white', col.lab = 'white', col.main = 'white', col.sub = 'white')

require(scales)
xl <- range(do.call(rbind, curves)$Date)
yl <- range(sapply(curves,function(x){x$elo}))
plot(xl, yl, col = 'transparent',
     xlab = 'Game No.', ylab='Elo Rating', main = 'Expansion Team Elo Ratings', las = 1)
abline(h = c(1300,1500), lty = 2)

for(i in 1:length(curves)){
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color1[i], lwd=4)
    lines(curves[[i]]$Date, curves[[i]]$elo, col=expanYrs$color2[i], lwd=1)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color2[i], pch=16, cex=2)
    points(curves[[i]]$Date[nrow(curves[[i]])], curves[[i]]$elo[nrow(curves[[i]])], col=expanYrs$color1[i], pch=16, cex = 1.5)
}

par(mar = c(5,0,4,1)+.1)
plot.new()
legend('left', legend = rep('', nrow(lgnd)), bty='n', lwd=4.5, 
       col = lgnd$color1, cex = 1.3)
legend('left', legend = lgnd$label, bty='n', lwd=1.1, 
       col = lgnd$color2, cex = 1.3)
```

By this metric, the Valkyries' 2025 season seems pretty closely aligned with the 1998 Detroit Shock, who actually had a slightly higher rating through 30 games.
Although they entered the league at very different points in its history, these seem to be the two best showings by any expansion team.

