---
title: "WNBA Standings"
format: 
  html:
    fig-width: 10
    fig-height: 6
    toc: false
lightbox: true
---

```{r}
#| include: false
source('code/setupStandings.R')
source('code/setupElo.R')
```


::: {.panel-tabset}

## 2025

```{r}
#| echo: false
#| message: false
year <- 2025
sched <- getSchedule(year)
# standings <- getStandings(year)
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
```


## 2024

```{r}
#| echo: false
#| message: false
year <- 2024
sched <- readRDS(paste0('data/sched',year,'.rds'))
# standings <- getStandings(year)
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
makeWNBAeloGraph(year, allgames)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
makeWNBAeloGraph(year, allgames, mode = 'dark')
```


## 2023
```{r}
#| echo: false
#| message: false
year <- 2023
sched <- readRDS(paste0('data/sched',year,'.rds'))
#standings <- readRDS(paste0('data/standings',year,'.rds'))
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
makeWNBAeloGraph(year, allgames)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
makeWNBAeloGraph(year, allgames, mode = 'dark')
```

## 2022
```{r}
#| echo: false
#| message: false
year <- 2022
sched <- readRDS(paste0('data/sched',year,'.rds'))
#standings <- readRDS(paste0('data/standings',year,'.rds'))
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
makeWNBAeloGraph(year, allgames)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
makeWNBAeloGraph(year, allgames, mode = 'dark')
```

## 2021
```{r}
#| echo: false
#| message: false
year <- 2021
sched <- readRDS(paste0('data/sched',year,'.rds'))
#standings <- readRDS(paste0('data/standings',year,'.rds'))
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
makeWNBAeloGraph(year, allgames)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
makeWNBAeloGraph(year, allgames, mode = 'dark')
```

## 2020
```{r}
#| echo: false
#| message: false
year <- 2020
sched <- readRDS(paste0('data/sched',year,'.rds'))
#standings <- readRDS(paste0('data/standings',year,'.rds'))
```

```{r}
#| echo: false
#| message: false
#| classes: light-mode
makeWNBAstandingsGraph(year, sched)
makeWNBAeloGraph(year, allgames)
```

```{r}
#| echo: false
#| message: false
#| classes: dark-mode
makeWNBAstandingsGraph(year, sched, mode = 'dark')
makeWNBAeloGraph(year, allgames, mode = 'dark')
```


:::


### About these plots

Standings plots show how many games above or below .500 each team was throughout the course of the season. Ties are broken arbitrarily, as I didn't actually code up the tie-breaker scenarios.

Elo ratings are a way of estimating a team's "true" skill level, invented by Arpad Elo to rank chess players ([Wikipedia](https://en.wikipedia.org/wiki/Elo_rating_system)). The rating system used here was adapted from the WNBA ratings system implemented by fivethirtyeight.com. Ratings are partially carried over from year to year, which is why the lines have different starting values. The long-run average rating is 1500.
